---
title: "exercise"
author: "Barry Quinn"
date: "13/02/2022"
output: html_document
---

## Learning outcomes

We have introduced the following graphics functions:

  - `gglagplot`
  - `ggAcf`

Explore the following time series using these functions. Can you spot any seasonality, cyclicity and trend? What do you learn about the series?

  - `ni_hsales_ts`
  - `vix_ts`
  - `ftse_m_ts`

## Which is which?

```{r, fig.height=6, fig.width=12, echo=FALSE, warning=FALSE, out.width="11.5cm"}
tp1 <- autoplot(ni_hsales_ts) + xlab("") + ylab("Total Verified Sales") +
  ggtitle("1. NI house sales")
tp2 <- autoplot(vix_ts) + xlab("") + ylab("") +
  ggtitle("2. VIX prices")
tp3 <- autoplot(ftse_m_ts) + xlab("") + ylab("") +
  ggtitle("3. FTSE 100 log returns")
acfa <- ggAcf(ni_hsales_ts, ci=0) + xlab("") + ggtitle("A") + ylim(-0.4,1)
acfb <- ggAcf(vix_ts, ci=0) + xlab("") + ggtitle("B") + ylim(-0.4,1)
acfc <- ggAcf(ftse_m_ts, ci=0) + xlab("") + ggtitle("C") + ylim(-0.4,1)
gridExtra::grid.arrange(tp1,tp2,tp3,
                        acfc,acfb,acfa,nrow=2)
```

## Your turn
  - load `fin7028_dat.RData` from canvas

We have introduced the following graphics functions:

  - `gglagplot`
  - `ggAcf`

Explore the following time series using these functions. Can you spot any seasonality, cyclicity and trend? What do you learn about the series?

  - `ni_hsales_ts`
  - `vix_ts`
  - `ftse_ts`

## Which is which?

```{r, fig.height=6, fig.width=12, echo=FALSE, warning=FALSE, out.width="11.5cm"}
tp1 <- autoplot(ni_hsales_ts) + xlab("") + ylab("Total Verified Sales") +
  ggtitle("1. NI house sales")
tp2 <- autoplot(vix_ts) + xlab("") + ylab("") +
  ggtitle("2. VIX prices")
tp3 <- autoplot(ftse_m_ts) + xlab("") + ylab("") +
  ggtitle("3. FTSE 100 log returns")
acfa <- ggAcf(ni_hsales_ts, ci=0) + xlab("") + ggtitle("A") + ylim(-0.4,1)
acfb <- ggAcf(vix_ts, ci=0) + xlab("") + ggtitle("B") + ylim(-0.4,1)
acfc <- ggAcf(ftse_m_ts, ci=0) + xlab("") + ggtitle("C") + ylim(-0.4,1)
gridExtra::grid.arrange(tp1,tp2,tp3,
                        acfc,acfb,acfa,nrow=2)
```

## Which is which?

1. NI house sales is A ACF plot
2. VIX prices is B ACF plot
3. FTSE 100 log returns is C ACF plot
